---
title: "RCT_comparison"
format: html
editor: visual
---

# Comparison of Methods for Analyzing RCT Data Regarding Baseline Treatment Group Differences

Motivation:

-   Seeing multiple approaches for analyzing fairly basic structures of RCT data
-   Reading that some strategies better adjust for **regression to the mean** because of baseline differences in treatment and control group (Twisk, 2018)
-   Wanting to understand the DAG structures that illustrate the "difference in differences" and "longitudinal analysis of covariance" approaches/implied data generating processes

```{r}
#| echo: false
#| output: false
# setup
rm(list=ls())
library(data.table)
library(ggplot2)
library(ggdag)
library(dagitty)
library(gridExtra)

## define useful functions
expit <- function(x) {exp(x)/(1+exp(x))}

## small number to have more likelihood for imbalance
N <- 30
niter <- 1:1000
```

## Analysis

### Overview
(1) Simulated RCT data. The treatment, $X$, was binary, and the outcome, $Y$, was simulated at baseline and after treatment. RCT data were simulated using multiple data generating processes.

(2) Ran models to estimate the effect of treatment on the outcome of interest, comparing error over many simulations, as well as examining relationship between error and the baseline imbalance in the outcome.

### Data Generating Processes

I tested multiple data generating processes to understand whether differences in the estimated effects using different methods:

(1) $Y_t=\beta_0 + \beta_X X + \beta_{Y_0} Y_0 + \epsilon$

(2) $Y_t=\beta_0 + \beta_X X + \beta_{Y_0} Y_0 + \beta_{X*Y_0} X*Y_0 + \epsilon$

```{r}
#| echo: false

dag1 <- dagitty::dagitty(
'dag {
  "E*Y0" [pos="0.1,0.357"]
  E [exposure,pos="-0.239,-0.459"]
  Y0 [pos="-0.129,1.076"]
  Y1 [outcome,pos="0.752,0.464"]
  E -> "E*Y0"
  E -> Y1
  Y0 -> "E*Y0"
  Y0 -> Y1
}'
)

dag2 <- dagitty::dagitty(
'dag {
  "E*Y0" [pos="0.1,0.357"]
  E [exposure,pos="-0.239,-0.459"]
  Y0 [pos="-0.129,1.076"]
  Y1 [outcome,pos="0.752,0.464"]
  "E*Y0" -> Y1
  E -> "E*Y0"
  E -> Y1
  Y0 -> "E*Y0"
  Y0 -> Y1
}'
)

gg1 <- ggdag(dag1, layout = "circle") + theme_dag_blank() + annotate("text",x=.45,y=.48,label="Effect = 0") + ggtitle("DGP 1: No Interaction\n(on Additive Scale)")

gg2 <- ggdag(dag2, layout = "circle") + theme_dag_blank() + 
  annotate("text",x=.45,y=.48,label="Effect =/= 0") + ggtitle("DGP 2: Interaction\n(on Additive Scale)")

grid.arrange(gg1, gg2, nrow = 1)

#ggplot(dag)


```

### Models Fit

I compared the following analytic approaches defined in Twisk et al.:

(1) "Longitudinal Analysis of Covariance"
(a) Unsaturated
$Y_t=\beta_0 + \beta_X X + \beta_{Y_0} Y_0 + \epsilon$
(b) Saturated
$Y_t=\beta_0 + \beta_X X + \beta_{Y_0} Y_0 + \beta_{X*Y_0} X*Y_0 + \epsilon$

(2) "Repeated Measures Analysis" (Difference in Differences)
$Y_t=\beta_0 + \beta_X X + \beta_{time} time + \beta_{time*X} time*X + \epsilon$





```{r}
#| echo: false

bd <- c()
m1 <- c()
m2 <- c()
for (i in niter) {
  cat(paste0(i)); flush.console()
  ## generate data
  tx <- sample(x=1:N,size=.5*N) ## make treatment randomization exactly 50%
  ## continuous first because it seems easier
  Y0 <- rnorm(n=N,mean=10,sd=1.5)
  E <- as.numeric(c(1:N) %in% tx)
  Y1 <- (1*Y0 + E*(-2) + rnorm(n=N,mean=0,sd=.2))
  id=1:N
  d <- data.table(as.data.frame(cbind(Y0,E,Y1,id)))
  
  ## data formatted for DiD
  d2 <- melt(copy(d),id.vars=c("id","E"),variable.name="time",value.name="Y")
  d2[,time:=as.numeric(substr(time,2,3))]
  d2[,id:=factor(id)]
  
  ## find imbalance in baseline values
  bd[i] <- mean(d[E==1]$Y0)-mean(d[E==0]$Y0)
  
  ## regression models
  mod1 <- lm(data=d2,formula="Y ~ E + time + E*time")
  # did2 <- lmer(data=d2,formula="Y ~ E + time + E*time + (1 | id)") in this case exactly the same
  mod2 <- lm(data=d,formula="Y1 ~ Y0 + E")
  m1[i] <- summary(mod1)$coefficients["E:time","Estimate"]
  m2[i] <- summary(mod2)$coefficients["E","Estimate"]
    
}

m1diff <- abs(m1+2)
m2diff <- abs(m2+2)
mean(m1)
mean(m2)
summary(lm(m1diff ~ bd))
summary(lm(m2diff ~ bd))
mean(bd)

```

Notes:

Note to look back: https://diff.healthpolicydatascience.org/#regression

not the same but maybe related https://www.sciencedirect.com/science/article/abs/pii/S0304407614002437?via%3Dihub https://onlinelibrary.wiley.com/doi/10.1111/1475-6773.13017 https://onlinelibrary.wiley.com/doi/full/10.1111/1475-6773.13015 https://pubmed.ncbi.nlm.nih.gov/33978956/

Note: ATT/ATE?

Note: Do with continuous X too?
